---
title: "BINF-6110-Assignment4"
author: "Kexin Gong"
date: "`r Sys.Date()`"
output: html_document
---

0. Load packages
```{r setup, message=FALSE, warning=FALSE}
# Seurat: core single-cell workflow
# dplyr/tibble: data wrangling
# ggplot2/patchwork: plotting and figure layout
# clusterProfiler + org.Mm.eg.db: enrichment analysis
# future: avoid parallel-related Seurat issues on large objects
# pheatmap: optional heatmap utility

library(Seurat)
library(dplyr)
library(ggplot2)
library(patchwork)
library(clusterProfiler)
library(org.Mm.eg.db)
library(tibble)
library(future)
library(pheatmap)

# Use sequential processing to avoid future/parallel memory errors
plan("sequential")
options(future.globals.maxSize = 8 * 1024^3)
options(future.fork.enable = FALSE)
```

1. Import data
```{r import-data}
# Read the Seurat object provided for the assignment.
data <- readRDS("../data/seurat_ass4.rds")
```


2. QC visualization
```{r qc-violin, fig.width=12, fig.height=5}
# Check whether sequencing depth and detected gene numbers are broadly comparable across time points before downstream analysis.
# We are visualizing QC only and not applying extra filtering here, because the data already looked clean in earlier inspection.

# Violin plot for number of detected genes per cell
p1 <- VlnPlot(
  data,
  features = "nFeature_RNA",
  group.by = "time",
  pt.size = 0
) +
  geom_boxplot(width = 0.1, fill = "white") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 14)

# Violin plot for UMI counts per cell
p2 <- VlnPlot(
  data,
  features = "nCount_RNA",
  group.by = "time",
  pt.size = 0
) +
  geom_boxplot(width = 0.1, fill = "white") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 14)

print(p1)
print(p2)

ggsave("../figures/1.QC_nFeature.png", p1, width = 8, height = 5, dpi = 300)
ggsave("../figures/2.QC_nCount.png", p2, width = 8, height = 5, dpi = 300)
```

```{r qc-scatter, fig.width=7, fig.height=5}
# Scatter plot showing the relationship between UMI counts and detected genes. 
# A strong positive correlation indicates good library complexity and overall healthy QC structure.
p_scatter <- FeatureScatter(
  data,
  feature1 = "nCount_RNA",
  feature2 = "nFeature_RNA",
  group.by = "time"
)

print(p_scatter)
ggsave("../figures/3.QC_scatter.png", p_scatter, width = 6, height = 5, dpi = 300)

# Explicitly calculate the correlation used in the QC description
cor_value <- cor(data$nCount_RNA, data$nFeature_RNA)
cor_value
```

3. Focus on respiratory mucosa (RM) for downstream analysis
```{r rm subset}
# Restrict to RM because:
# 1) RM is the main region relevant to the infection story
# 2) downstream marker, composition, and remodeling analyses are easier to support in this subset
data_rm <- subset(data, subset = organ_custom == "RM")

# Check that all time points are represented in RM
table(data_rm$time)
```

```{r rm-clustering}
DefaultAssay(data_rm) <- "RNA"

data_rm <- NormalizeData(data_rm)
data_rm <- FindVariableFeatures(data_rm, selection.method = "vst", nfeatures = 2000)
data_rm <- ScaleData(data_rm)
data_rm <- RunPCA(data_rm, verbose = FALSE)
data_rm <- RunUMAP(data_rm, dims = 1:20)
data_rm <- FindNeighbors(data_rm, dims = 1:20)
data_rm <- FindClusters(data_rm, resolution = 0.5)
```

```{r rm-umap-cluster, fig.width=8, fig.height=6}
# UMAP colored by cluster identity
p_rm_cluster <- DimPlot(
  data_rm,
  reduction = "umap",
  group.by = "seurat_clusters",
  label = TRUE,
  raster = FALSE
) + theme_minimal()

print(p_rm_cluster)
ggsave("../figures/4.UMAP_clusters.png", p_rm_cluster, width = 7, height = 6, dpi = 300)
```

4. Annotation (manual, based on canonical markers)
```{r marker-violin, fig.width=10, fig.height=8}
# 1) use known canonical markers to identify broad cell types
# 2) use FindAllMarkers on RM only for support, not as the sole basis of annotation
# This avoids over-interpreting noisy cluster-specific genes while still giving biologically grounded labels

# Marker violin plot for broad annotation
plot_markers <- VlnPlot(
  data_rm,
  features = c("Epcam", "Cd3e", "Lyz2", "Krt13", "Cnga4"),
  group.by = "seurat_clusters",
  stack = TRUE,
  flip = TRUE
) + theme(legend.position = "none")

print(plot_markers)
ggsave("../figures/5.Marker_violin.png", plot_markers, width = 9, height = 7, dpi = 300)
```

```{r find-rm-markers}
# Set cluster identity before marker analysis
Idents(data_rm) <- "seurat_clusters"

# Find cluster markers within RM
# max.cells.per.ident is used to reduce memory pressure
markers_rm <- FindAllMarkers(
  data_rm,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25,
  max.cells.per.ident = 2000
)

# Extract top markers per cluster for inspection
top_markers_rm <- markers_rm %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 10) %>%
  ungroup()

print(top_markers_rm)
```

```{r manual-annotation}
# Manual annotation derived from: marker violin plots, top cluster markers and biological interpretation of the RM dataset
data_rm$celltype <- "Other"

# Olfactory neuron-like clusters
data_rm$celltype[data_rm$seurat_clusters %in% c(0, 29)] <- "Olfactory_neuron"

# T-cell clusters
data_rm$celltype[data_rm$seurat_clusters %in% c(2, 28)] <- "T_cells"

# Myeloid clusters
data_rm$celltype[data_rm$seurat_clusters %in% c(3, 4, 8)] <- "Myeloid"

# IFN-responsive epithelial-like clusters
data_rm$celltype[data_rm$seurat_clusters %in% c(5, 9, 10)] <- "IFN_epithelial"

# Remodeling epithelial cluster of main interest
data_rm$celltype[data_rm$seurat_clusters %in% c(24)] <- "Remodeling_epithelial"

# Remaining epithelial-enriched clusters
data_rm$celltype[data_rm$seurat_clusters %in% c(
  1, 6, 7, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 25, 26, 27
)] <- "Epithelial"
```

```{r annotated-umap, fig.width=10, fig.height=7}
# Annotated UMAP
plot_annoted <- DimPlot(
  data_rm,
  reduction = "umap",
  group.by = "celltype",
  label = TRUE,
  repel = TRUE,
  raster = FALSE
) + theme_minimal()

print(plot_annoted)
ggsave("../figures/6.UMAP_annotation.png", plot_annoted, width = 8, height = 6, dpi = 300)
```

5. Feature plots of genes of interest

5A. Canonical markers supporting annotation
```{r feature-annotation, fig.width=10, fig.height=8}
# validate annotation
genes_annotation <- c("Epcam", "Cd3e", "Lyz2", "Krt13", "Cnga4")
genes_annotation <- intersect(genes_annotation, rownames(data_rm))

p_annot_features <- FeaturePlot(
  data_rm,
  features = genes_annotation,
  keep.scale = "all",
  raster = FALSE
)

print(p_annot_features)
ggsave("../figures/7.Feature_annotation.png", p_annot_features, width = 10, height = 8, dpi = 300)
```

5B. Genes supporting the biological story
```{r feature-story, fig.width=10, fig.height=8}
# support the biological story of IFN response and remodeling
genes_story <- c("Isg15", "Ifit1", "Rsad2", "Cxcl16", "Cd274")
genes_story <- intersect(genes_story, rownames(data_rm))

p_story_features <- FeaturePlot(
  data_rm,
  features = genes_story,
  keep.scale = "all",
  raster = FALSE
)

print(p_story_features)
ggsave("../figures/8.Feature_story.png", p_story_features, width = 10, height = 8, dpi = 300)
```

6. Differential expression analysis of one cluster
```{r}
# Cluster 24 is treated as a remodeling epithelial population.
# For this cluster, the most informative comparison is not D14 vs Naive within the cluster, but rather cluster 24 versus all other RM cells.
# This identifies genes that define the remodeling epithelial state.

# cluster 24 = remodeling epithelial
cluster_of_interest <- 24
```

6A. Check abundance of cluster 24 across time
```{r cluster24-abundance}
# This captures the temporal emergence/enrichment of cluster 24
cluster_abundance <- data_rm@meta.data %>%
  mutate(cluster_flag = ifelse(seurat_clusters == cluster_of_interest, "cluster24", "other")) %>%
  group_by(time, cluster_flag) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(time) %>%
  mutate(freq = n / sum(n))

print(cluster_abundance)
```

```{r cluster24-abundance-plot, fig.width=7, fig.height=8}
# Plot relative abundance of cluster 24 across time
p_cluster24_abundance <- ggplot(
  cluster_abundance %>% filter(cluster_flag == "cluster24"),
  aes(x = time, y = freq, group = 1)
) +
  geom_point(size = 3, color = "firebrick") +
  geom_line(color = "firebrick") +
  theme_minimal() +
  labs(
    x = "Time",
    y = "Proportion"
  )

print(p_cluster24_abundance)
ggsave("../figures/9.Cluster24_abundance.png", p_cluster24_abundance, width = 6, height = 4.5, dpi = 300)
```

6B. Identify marker genes of cluster 24 versus all other RM cells
```{r cluster24-de}
Idents(data_rm) <- "seurat_clusters"

de_cluster <- FindMarkers(
  data_rm,
  ident.1 = as.character(cluster_of_interest),
  only.pos = FALSE,
  logfc.threshold = 0.25,
  min.pct = 0.25,
  max.cells.per.ident = 2000
)

de_cluster <- de_cluster %>%
  rownames_to_column("gene") %>%
  arrange(p_val_adj, desc(avg_log2FC))

print(head(de_cluster, 20))
```


7. ORA / GSEA for cluster 24 markers

7A. ORA using significantly upregulated genes in cluster 24
```{r ora-cluster24, fig.width=10, fig.height=6}
# summarize significantly upregulated genes

sig_up_genes <- de_cluster %>%
  filter(p_val_adj < 0.05, avg_log2FC > 0) %>%
  pull(gene) %>%
  intersect(keys(org.Mm.eg.db, keytype = "SYMBOL"))

if (length(sig_up_genes) == 0) {
  message("No significant upregulated genes found for ORA.")
} else {
  
  ora_res <- enrichGO(
    gene = sig_up_genes,
    OrgDb = org.Mm.eg.db,
    keyType = "SYMBOL",
    ont = "BP",
    pAdjustMethod = "BH"
  )
  
  ora_df <- as.data.frame(ora_res)
  print(head(ora_df, 20))

  p_ora <- dotplot(ora_res, showCategory = 12) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.y = element_text(size = 10),
      plot.title = element_text(hjust = 0.5)
    ) +
    labs(
      title = "GO enrichment of cluster 24 markers",
      x = "Gene Ratio",
      y = NULL
    ) +
    scale_y_discrete(labels = function(x) stringr::str_wrap(x, width = 30))
  
  print(p_ora)
  ggsave("../figures/10.ORA_dotplot.png", p_ora, width = 9, height = 5.5, dpi = 300)
}
```

7B. GSEA using ranked cluster-specific marker list
```{r gsea-cluster24, fig.width=10, fig.height=6}
# uses the full ranked list and can capture broader programs
gene_list <- de_cluster$avg_log2FC
names(gene_list) <- de_cluster$gene
gene_list <- gene_list[!is.na(gene_list)]
gene_list <- sort(gene_list, decreasing = TRUE)
gene_list <- gene_list[names(gene_list) %in% keys(org.Mm.eg.db, keytype = "SYMBOL")]

if (length(gene_list) <= 10) {
  message("Not enough genes available for GSEA.")
} else {
  gsea_res <- gseGO(
    geneList = gene_list,
    OrgDb = org.Mm.eg.db,
    keyType = "SYMBOL",
    ont = "BP",
    pvalueCutoff = 0.05
  )

  gsea_df <- as.data.frame(gsea_res)
  print(head(gsea_df, 20))

  if (nrow(gsea_df) == 0) {
    message("No significant GSEA terms found.")
  } else {
    p_gsea <- dotplot(gsea_res, showCategory = 12) +
      theme_minimal(base_size = 12) +
      theme(
        axis.text.y = element_text(size = 10),
        plot.title = element_text(hjust = 0.5)
      ) +
      labs(
        x = "Gene Ratio",
        y = NULL
      ) +
      scale_y_discrete(labels = function(x) stringr::str_wrap(x, width = 30))

    print(p_gsea)
    ggsave("../figures/11.Cluster24_GSEA_dotplot.png", p_gsea, width = 10, height = 6, dpi = 300)
  }
}
```

8. Additional visualization: cell composition across time
```{r cell-composition, fig.width=10, fig.height=6}
#cluster 24 abundance plot with a broader view of time-dependent
#shifts in annotated cell types

celltype_prop <- data_rm@meta.data %>%
  group_by(time, celltype) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(time) %>%
  mutate(freq = n / sum(n))

p_comp <- ggplot(celltype_prop, aes(x = time, y = freq, fill = celltype)) +
  geom_bar(stat = "identity") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Cell type composition across time",
    x = "Time",
    y = "Proportion"
  ) +
  scale_fill_brewer(palette = "Set2")

print(p_comp)
ggsave("../figures/12.Celltype_composition.png", p_comp, width = 8, height = 5.5, dpi = 300)
```

9. Additional visualization: marker DotPlot across clusters
```{r marker dot-plot, fig.width=10, fig.height=7}
# summarizes how representative cluster 24 marker genes are distributed across all RM clusters
Idents(data_rm) <- "seurat_clusters"

marker_panel <- c("Krt13", "Krt6b", "Plet1", "Csta1", "Prss27", "Calml3", "Pglyrp4")
marker_panel <- intersect(marker_panel, rownames(data_rm))

p_dot <- DotPlot(
  data_rm,
  features = marker_panel,
  group.by = "seurat_clusters",
  cols = c("grey90", "mediumpurple2", "blue3")
) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Expression of representative cluster 24 marker genes",
    x = "Genes",
    y = "Cluster"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    panel.grid.major = element_line(color = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )

print(p_dot)
ggsave("../figures/13.Marker_dotplot.png", p_dot, width = 9, height = 6, dpi = 300)
```
